---
title: "Lesson I. R language Programming Environment"
author: "Ruth Gómez Graciani"
date: "2017/10/16"
output:
 pdf_document: default
 html_document: default
---

#### 1 - Create	a	vector	called	v	and	assign	the	values	(3.14,	123,	0.01)

```{r comment=''}
v <- c(3.14, 123,	0.01)
v
```
 
#### 2- Create	a	new	vector	that	stores	three	values:	v,	123	and	v

```{r comment=''}
a <- c(v, 123, v)
a
```

#### 3- Calculate	the	square	root	of	vector	v	and	store	it	in	a	new	varieble

```{r comment=''}
my_sqrt <- sqrt(v)
my_sqrt
```

#### 4- Now	calculate	the	value	of	v	divided	by	my_sqrt

```{r comment=''}
v / my_sqrt
```

#### 5- Add	this	two	vectors:	(1,	2,	3,	4)	and	(0,100).	What	do	you	see?	Can	two	vectors	of	different	length	be	added?

```{r comment=''}
c(1, 2, 3, 4) + c(0, 100)
```

Yes, they can, because the largest vector is a multiple from the smaller one. The small vector is repeated twice in order to add it to the first one.

#### 6- Create	a	sequence	of	numbers	from	1	to	19	using	the	:	operator

```{r comment=''}
1:19
```

#### 7- Now	use	seq()	to	create	a	sequence	from	1	to	10	incrementing	by	0.5

```{r comment=''}
seq(1, 10, 0.5)
```

#### 8- Use	seq()	to	generate	a	list	of	50	numbers	between	50	and	999	and	store the	list	in n

```{r comment=''}
n <- seq(50, 999, length.out = 50)
n
```

#### 9- Apply	the	length()	function	to	the	variable	n.	Which is	the	result?

```{r comment=''}
length(n)
```

The result is the number of elements inside the vector

#### 10- Now	generate	a	new	list	of	numbers	starting	from	1	that	has	the	same length	as	n	variable

```{r comment=''}
b <- list(1:length(n))
b
```

#### 11- Create	a	vector	with	100	zeros

```{r comment=''}
c <- rep(0, 100)
c
```

#### 12- Create	a	new	vector	with	10	repetitions	of	the	vector	("a",	"b",	"c")

```{r comment=''}
d <- rep(c('a', 'b', 'c'), 10)
d
```
 
#### 13- Now	create	a	new	vector	that	contains	10	a	letters,	then	10	b's,	then	10 c's	and	save	it	into	a	variable	called	abc

```{r comment=''}
abc <- c(rep('a', 10), rep('b', 10), rep('c', 10))
abc
```

#### 14- Create	a	vector	with	the	values	(1,100,-25,365)	and	save	it	into	v1	variable

```{r comment=''}
v1 <- c(1, 100, -25, 365)
v1
```

#### 15- now	create	a	new	variable	bool	that	gets	the	result	of	v1<1.	What	is	the result	of this	operation?	What	kind	of	value	you	see?

```{r comment=''}
bool <- v1 < 1 
bool
```

The result of the operation is TRUE when the inequality is accomplished, and FALSE when not, for each value of the vector. The value is a logical operator, as we can see when the `str()` function is used:

```{r comment=''}
str(bool)
```

#### 16- Now	take	abc	variable	from	question	13	and	generate	a	single	string	 value.	That	is,	something	like:	(a,b,c)	->	"abc"

```{r comment=''}
abc <- paste(abc, collapse="")
abc
```
 
#### 17- Create	a	new	matrix	with	4	rows	and	5	columns	from	a	new	vector	that	contains	the	values	from 1	to	20.	Store	the	results	in	variable	m1

```{r comment=''}
e <- c(1:20)
m1 <- matrix(e, 4,5)
m1
```

#### 18- Now	create	a	vector	called	students	with	the	values ("Ana","John","Pedro","Joan")	and	create	a	new	matrix	adding	students	as a	new	column	to	m1	matrix.	What	happens	to	numeric	values?

```{r comment=''}
names <- c("Ana", "John", "Pedro", "Joan")
m2 <- matrix(c(m1, names), 4, 6)
m2
```

The numerical values have been converted into characters

#### 19- Now	create	a	new	data.frame()	called	df	with	students	vector	and	m1 matrix.	What	difference	do	you	see	with	the	previous	matrix?

```{r comment=''}
df <- data.frame(m1, names)
df
```

The data frame can contain different value types

#### 20- create	a	new	vector	with	the	labels	"name",	"age", "weight",	"bp",	 "rate",	"test".	Now	use	colnames()	to	set	the	colnames	attribute of	our	data	frame	df.	Show	your	final	data	frame 

```{r comment=''}
labels <- c("name",	"age", "weight",	"bp",	 "rate",	"test")
colnames(df) <- labels
df
```
 
#### 21- Calculate	the	mean	of	a	vector	with	values	(10,11,12)

```{r comment=''}
f <- c(10, 11, 12)
mean(f)
```

#### 22- Now	create	a	new	function	call	new_mean	that	receives	a	vector	as	a	 parameter,	then	calculates	the	sum	of	the	vector,	the	length	of	the	vector and	return	the	mean	of	the	vector.

```{r comment=''}
new_mean <- function(vector) {
  summatory <- sum(vector)
  numofels <- length(vector)
  return(summatory / numofels)
}

new_mean(f)
```

#### 23- Use	sample()	function	to	simulate	a	rolling	dice.	That	is,	randomly	select four	numbers	between	1	and	6	with	replacement.

```{r comment=''}
sample(1:6, 4, replace = T)
```

#### 24- Simulate	50	flips	of	an	unfair	two-sided	coin.	The	probabilities	here are 0.7	for	heads	and	0.3	for	tails.	Use	sample() to	draw	a	sample	of	size	50.	 Store	the	results	in	variable	flip.	How	would	you	test	if	the	results	follow the probability	defined?
```{r comment=''}
coin <- c('heads', 'tails')
prob <- c(0.7, 0.3)
flip <- sample(coin, 50, prob, replace = T)
table(flip) / 50
```

The `table` function counts the number of elements inside the vector, and dividing it by the total number of elements we obtain the proportion or probability. 

#### 25- How	would	you	generate	10	random	numbers	from	a	standard	normal distribution	with	a	mean of	150	and	a	standard	deviation	of	10?

```{r comment=''}
rnorm(10, 150, 10)
```

#### 26- Now,	how	would	you	generate	100	sets	of	random	numbers	like	the previous	question?	Save	results	in	a	variable	called	rp

```{r comment=''}
rp <- matrix(nrow = 10, ncol = 100)

for(i in c(1:100)){
  rp[,i] <- rnorm(10, 150, 10)
 }
```

The variable `rp` is partially below:

```{r comment=''}
rp[, 1:7]
```

#### 27- Use	colMeans()	function	to	find	the	mean	of	each	column	of	rp	and	plot a	histogram	

```{r comment=''}
g <- colMeans(rp)
g
```

```{r, echo = FALSE, warning= FALSE }
library(ggplot2)
```

```{r comment=''}

ggplot()+
 aes(g)+ 
 geom_histogram(binwidth=1, colour="#413839", fill="white")+
 labs(title="Mean values histogram", x = " Mean values in rp", y = "Count" )+
 theme(plot.title = element_text(hjust = 0.5))


```

#### 28- Create	a	vector	X	with	10	values	and	calculate	the	cumulative	sums	for	this	vector	using	a	for	loop.	

```{r comment=''}
X <- sample(100, 10)
X
```

```{r comment=''}

for (i in c(2:length(X))) {
  X[i] <- X[i] + X[i - 1]
}
X
```

#### 29- Create	two	vectors,	Z1	and	Z2.	Z1	is	a	vector	of	20	elements	filled	with	20	random	numbers	from	a	standard	normal	distribution.	Z2	is	a	string	of 20	zeros.	Design	a	for	loop	that	counts	from	1	to	20.	For	each	value,	it	has	to	evaluate	if	the	value	is	greater	than	0.	For	all	those	values,	it	must	write	a	1	in	the	same	position	of	the	vector	Z2.	For	lower	than	0	values	it	must write	a	-1.	Is	it	possible	to	do	the	same	thing	using	which()? How	would	you	 do	it?

With the `for` loop
```{r comment=''}

Z1 <- rnorm(20)
Z2 <- rep(0, 20)
for (i in c(1:20)) {
  if (Z1[i] > 0) {
    Z2[i] <- 1
  } else{
    Z2[i] <- -1
  }
}
```

With `which()`
```{r comment=''}

Z2 <- rep(0,20)
Z2[which(Z1 > 0)] <- 1
Z2[which(Z1 < 0)] <- -1
```

#### 30- Now	we	are	going	to	import	a	data	file:plants.csv	and	use	read.table()function	to	store	the	new	table	in	an	object	called	plants

```{r comment=''}
plants <- read.table("Data/plants2017_10_13D19_4_42 (1).csv", sep = ",")
```

#### 31- How	do	you	find	the	rows	and	columns	of	the	variable	plants?	And	the size	in	bytes?

As we can see below, there is a large number of rows:

```{r comment=''}
nrow(plants)
ncol(plants)
```

A small subset of the row names will be displayed. All the column names will be displayed

```{r comment=''}
rownames(plants)[1:10]
colnames(plants)
```

The size in bytes:

```{r comment=''}
object.size(plants)
```

#### 32- How	would	you	see	the	first	rows	of	the	experiment?

```{r comment=''}
head(plants)
```
 
#### 33- What	information	is	summary()	providing	when	applied	to	the	experiment	variable?	What	NA	plants$Precip_Minmeans?	What	information	is	str()	providing?

`summary()` shows basic statistical values of the vector. In this case, I have assumed that one of the experiment variables is pH_Min.

```{r comment=''}
summary(plants$pH_Min)
```

NA states fot Not Available (it is missing data)

`str()` provides the type of variable and the elements that can be found inside it. As an example, we can see that str is a datagfarme and therefore it can contain different types of values. 

```{r comment=''}
str(plants, strict.width="cut")
```

#### 34- Create	a	function	to	calculate	the	minimum	Temp_Min_F	and	the	average	Precip_Min	of	the	table	plants

There are already functions to calculate the minimum and average values of a vector, I have assumed that the function has to retrieve both at the same time, from the columns specified, when a datafarme is provided. 

```{r comment=''}
Calculate_plants <- function(plants){
  minimum <- min(plants$Temp_Min_F, na.rm = TRUE)
  average <- mean(plants$Precip_Min, na.rm = TRUE)
  return(c(minimum, average))
}

Calculate_plants(plants)

```

### To	be	evaluated:
#### Create	a	function	that:
* Creates	a	20	by	20	matrix	of	0	values
* Places	in	a	random	position	inside	the	matrix	one	vector	of	size	6	and	 represents	it	with	six	consecutive	"1"	values.	This	vector	will	be	our	 ship
* Receives	two	vectors	of	six	values	as	inputs.	X	and	Y.	Both	vectors	store	six	values	between	1	and	20	and	represent	coordinates	in	the	 matrix
* The	function	returns	a	vector	with	six	values	depending	if	the	coordinates	have	hit	or	miss	our	ship. Example=("hit",	"miss",	"miss", "miss",	"miss",	"hit")

#### Make	the	number	of	ships	in	the	matrix	a	parameter	given	by	the	user and	export	the	matrix	to	a	text	file.

The ships can be vertical or horizontal, but they can't overlap.

```{r comment=''}
Battleship <- function(X, Y, ships) {
  # Create the empty matrix
  m1 <- matrix(0, 20, 20)
  # Iterate over ships
  for (s in c(1:ships)) {
    # We avoid solapping ships
    done <- F
    while (done == F) {
      # Place in a random position, and random direction
      variablenum <- c(1:6) + sample(c(1:14), 1)
      fixnum <- sample(c(1:20), 1)
      random <- sample(c(1:2))
      order <- list()
      order[[random[1]]] <- variablenum
      order[[random[2]]] <- fixnum
      if (sum(m1[order[[1]], order[[2]]]) == 0) {
        m1[order[[1]], order[[2]]] <- c(rep(1, 6))
        done <- T
      }
    }
  }
  # Creates result
  result <- as.factor(diag(m1[X, Y]))
  levels(result) <- c('miss', 'hit')
  # Export matrix and return result
  write.table(m1, 'matrix.txt', col.names = F,row.names = F,quote = F )
  return(as.character(result))
}
```

Here there is an example of the result of the function:

```{r comment=''}
Battleship(c(1, 4, 15, 6, 17, 11), c(10, 14, 16, 2, 7, 20), 3)
```

The generated file looks like this (I am using `engine = 'bash'`):

```{r engine='bash', comment=''}
cat matrix.txt
```
